VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-17 doc/publications/journal2014/paper.tex
updates

To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib. 
In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
In section two we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
In section three we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
In the last section we discuss the pipeline for visualising the resulting cluster.

@@

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
Even the state-of-the-art tool, {\sc Admixture} does returns a low ARI=0.25.
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
Including AMR individuals, we observe that the majority of these individuals are placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Hunter2014}.

\subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}

Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.

We run this comparison on our in-house Hadoop cluster using 40 cores.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
%This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
% DENIS: This conflicts with the 30 seconds gain over {\sc ADAM} which you attribute to our clustering using sparse vectors,
% AIDAN: Only centroids are dense. Reworded to explain this
% DENIS: That means you allow more clusters ?
The increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
%AIDAN: not really.. 
%DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
%TODO A citation would be good here
We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.

@@

This highlights the issue for clinical application where ancestry influences treatment (e.g. HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity. 
% DENIS: statement about why we cannot do more with this data...
Unfortunately, this hypothesis could no not be tested as whole genome information is not available for these individuals.

\subsection*{Graph visualization}
%% Add the image here \ldots
A visualization of the clusters generated by \variantSpark{} from 1000 Genomes Project data, generated using Gephi \cite{ICWSM09154}, is 
shown in Figure 3. Three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
consists predominantly of individuals labelled EUR and AMR.

@@

\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 Million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.

@@

\subsection*{Datasets}
We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
We also used a subset of this dataset, the variants from chromosome 22, and this contains 494,328 alleles.
These individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).

@@

\subsection*{ADMIXTURE Implementation}
We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.

@@

\section*{Author's contributions}
    Text for this section \ldots

@@

\begin{figure}[h!]
\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{chr22-k4-gephi}
  \caption{\csentence{Visualization of \variantSpark{} clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
      \end{figure}

>>>>

To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib. 
In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
In the second section we discuss the pipeline for visualising the resulting cluster.
In section three we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
In the last section we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.

@@

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
Even the state-of-the-art tool, {\sc Admixture}, does return a low ARI=0.25. It should be noted that {\sc Admixture}'s underlying statistical model does not take linkage disequilibrium (LD) into account. 
Therefore removing variants with high LD may result in higher accuracy.   
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
This is due to the majority of AMR individuals being placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Pugach2015}.

\subsection*{Graph visualisation}
\variantSpark{} also supports visualisation of the resulting population clusters by generating a file that can be loaded in Gephi \cite{ICWSM09154}.
Figure 3 visualises the clusters generated by \variantSpark{} from the 1000 Genomes Project data. 
As discussed in the previous section, three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
consists predominantly of individuals labelled EUR and AMR providing a visual representation of the ARI of 0.84.

\subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.

We run this comparison on our in-house Hadoop cluster using 40 cores.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
%TODO: doubleckeck
The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries~\cite{Pugach2015}.

@@

This highlights the issue for clinical application where ancestry influences treatment (e.g. HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
As noted by Patterson et al ~\cite{Patterson2006}, more markers are needed for populations with low genetic divergence. 
It is hence likely that higher density genotyping (e.g. from genome sequencing) would help in elucidating population structure in this dataset, which demonstrates the need for fast whole-genome approaches. 
Unfortunately, this hypothesis can no not be tested as whole genome information is not available for these individuals.

@@

\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 Million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
This novel parallelisation approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.

@@

\subsection*{Datasets}
We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
We also used a subset of this dataset, the variants from chromosome 22, and this contains 494,328 alleles.
The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).

@@

\subsection*{{\sc Admixture} Implementation}
We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.

@@

\section*{Author's contributions}
    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NS obtained the PGP data and generated the graph.
    All authors contributed to the writing of the document.

@@

\begin{figure}[h!]
\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{chr22-k4-gephi-legend}
  \caption{\csentence{Visualization of \variantSpark{} clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
      \end{figure}
